MLH1 Results Outline

### 2020-02-04 doc/Main_Report_Results.rmd
filling in pvalues and making notes for the mouse mlh1 models

```{r mixed.model.data.setup, echo=FALSE, warning=FALSE}
# MOVE this into BIVDATA SETUP!!!

@@

#afriad this is removing NA's
DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel[( (DF.HetC.MixedModel$age.weeks < 25) &  (DF.HetC.MixedModel$age.days != 13) ) | is.na(DF.HetC.MixedModel$age.weeks), ] #keeps mice without ages, and removes juvinilles

@@

```{r MixedModel.1, echo=FALSE, digits = 3}
#THIS IS THE NEW MIXED MODEL

#reduced models
MLH1.MM.M1 <- lmer(mean_co ~ subsp * sex + (1|strain), data=DF.HetC.MixedModel.HQ) 
#summary(Reduced.strain.HQ)

drop1(MLH1.MM.M1, test="Chisq") # p = 0.0002045 ***

@@

as.data.frame(round(summary(MLH1.MM.M1)$tTable, 6))
#add a aption to this code chunk
#ranef(Reduced.strain.HQ)
```

@@

## Variance in CO count

DECIDE WHICH RESULTS TO DISPLAY.

in addition to the mean gwRR, the variance is also an important metric to describe

@@

meiotic Checkpoints for ensuring consistency (Hochwagen and Amon 2006, Li and Schimenti 2007, Jaramillo-Lambert 2010 and Kauppi 2013)

#### MLH1 variance Figures

Not sure why I have this section -- think this should be into a supplemental doc or something.

```{r variance.boxplots, echo=FALSE, eval=FALSE}
#how did I make this 
#Species.COcount.mixed <- lmer(nMLH1.foci ~ species + quality + (1|mouse), data=MLH1.hq.lab, REML = TRUE)
#Species.COcount.mixed.sum <- summary(Species.COcount.mixed)
#SpeciesCOcount.fix.est <- drop1(Species.COcount.mixed, test="Chisq") 
#SpeciesCOcount.Rand.est <- exactRLRT(Species.COcount.mixed)

#plot of variance
var.boxplot <- ggplot(DF.HetC.MixedModel.HQ, aes(x = as.factor(mouse), y = var, fill= strain,alpha=0.5))+ geom_point()+
 ggtitle("Boxplots of MLH1 distributions") + theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
 ylim( c((min(MLH1_data$adj_nMLH1.foci)-3) ), (max(MLH1_data$adj_nMLH1.foci)+3)) +   scale_fill_manual(values=colors_of_strains)
var.boxplot <- var.boxplot   + facet_wrap(sex ~ strain, scales="free")
#this plot shows variance as points/single values

#plot by sex / seperately
# HQ data DF.HetC.MixedModel.HQ

var.male <- ggplot(DF.HetC.MixedModel.HQ[DF.HetC.MixedModel.HQ$sex == "male",], aes(x = as.factor(mouse), y = var, fill= strain,alpha=0.5))+ geom_point()+
 ggtitle("Male variance levels") + theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  scale_fill_manual(values=colors_of_strains)+  ylim( c( 5, 20) )

var.male <- var.male   + facet_wrap( ~ strain, scales="free")

cV.male <- ggplot(DF.HetC.MixedModel.HQ[DF.HetC.MixedModel.HQ$sex == "male",], aes(x = as.factor(mouse), y = cV, fill= strain,alpha=0.5))+ geom_point()+
 ggtitle("Male cV levels") + theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  scale_fill_manual(values=colors_of_strains)+  ylim( c( 5, 20) )

cV.male <- cV.male   + facet_wrap( ~ strain, scales="free")

@@

```{r variance.hist, echo=FALSE, eval=FALSE}

#DF.HetC.MixedModel.HQ
Var.Hist <- ggplot(DF.HetC.MixedModel.HQ, aes(x = var, fill=sex)) + 
  geom_histogram(binwidth = 1, alpha=.6)  + 
  ggtitle("Histograms of within mouse variance, 10 cell min")
Var.Hist <- Var.Hist   + facet_wrap(~ strain, scales="free")
Var.Hist

Var.Hist.15cells <- ggplot(DF.HetC.MixedModel.HQ[DF.HetC.MixedModel.HQ$Ncells >14,], aes(x = var, fill=sex)) + 
  geom_histogram(binwidth = 1, alpha=.6)  +   ggtitle("Histograms of within mouse variance, 15 cell min")
Var.Hist.15cells <- Var.Hist.15cells   + facet_wrap(~ strain, scales="free")
Var.Hist.15cells

A general pattern is that the female data has higher within mouse variance (both var and cV). Some strains more than others. I think the dissection date needs to be tested as a predictor factor.  It might be tricky to add the dissection date metric. 
Test if dissection date is more significant predictor than strain (for females)
Also test how the number of 0CO bivalents per cell affects mouse with variance.

(LEW females with top 3 variance are all from different dissection dates)

I don't have enough mice with >15 cells for a complete idea of the distributions, but comparing single mice with enough good quality cells show be

@@

#DF.HetC.MixedModel.HQ
cV.Hist <- ggplot(DF.HetC.MixedModel.HQ, aes(x = cV, fill=sex)) + 
  geom_histogram(binwidth = 1, alpha=.6)  + 
  ggtitle("cV of within mouse variance, 10 cell min")
cV.Hist <- cV.Hist   + facet_wrap(~ strain, scales="free")
cV.Hist

@@

### Modeling Variance

2.	Reduced.strain.HQ <- lme(mean_co ~ subsp * sex, data=DF.HetC.MixedModel.HQ, random = ~1|strain)

3.	lm(mean_co ~ subsp * sex * strain, data=DF.HetC.MixedModel.HQ)

4.	strain.fixed <- lm(mean_co ~ sex * strain, data=DF.HetC.MixedModel.HQ)

$$mouse .var CO ~=~ subsp * sex + rand(strain*sex)+ \varepsilon$$

```{r M1.var, echo=FALSE}
#modeling within mouse variance
#use  DF.HetC.MixedModel.HQ

#HQ var.CO
M1.var_co <- lme( var ~ subsp * sex, data=DF.HetC.MixedModel.HQ, random=list(strain=pdDiag(~sex) ) )

#summary(HetC.MixedModel.HQ.var_co)

as.data.frame( round(summary(M1.var_co)$tTable, 6))

#ranef(HetC.MixedModel.HQ.var_co)

#CV
M1.cv_co <- lme( cV ~ subsp * sex, data=DF.HetC.MixedModel.HQ, random=list(strain=pdDiag(~sex) ) )

#summary(HetC.MixedModel.HQ.var_co)
as.data.frame( round(summary(M1.cv_co)$tTable, 6))
```

```{r M1.Q12_table, echo=FALSE}
#use the Q1Q2 calq'd table

#Q12_mouse_table_w.Ages$subsp <- ifelse(grepl(1, Q12_mouse_table_w.Ages$subsp), "Dom", 
#        ifelse(grepl(2, Q12_mouse_table_w.Ages$subsp), "Musc", Q12_mouse_table_w.Ages$subsp))

Q12_mouse_table_w.Ages <- add_subsp(Q12_mouse_table_w.Ages)

#need to change to chr first

Q12_mouse_table_w.Ages$subsp[Q12_mouse_table_w.Ages$strain== "MSM"] <- 'Mol'

Q12_mouse_table_w.Ages$subsp[Q12_mouse_table_w.Ages$strain== "MOLF"] <- "Mol"

M1.Q12_var <- lme( var ~ subsp * sex, data=Q12_mouse_table_w.Ages, random=list(strain=pdDiag(~sex) ) )

as.data.frame( round(summary(M1.Q12_var)$tTable, 6))

M1.Q12_CV <- lme( cV ~ subsp * sex, data=Q12_mouse_table_w.Ages, random=list(strain=pdDiag(~sex) ) )

as.data.frame( round(summary(M1.Q12_CV)$tTable, 6))

@@

$$mouse .cV CO ~=~ subsp * sex + rand(1|strain)+ \varepsilon$$

```{r M2.var, echo=FALSE}
#this model must be the same as M! -- since all the results are the same
M2.var <- lme(var ~ subsp * sex, data=DF.HetC.MixedModel.HQ, random = ~1|strain)

as.data.frame( round(summary(M2.var)$tTable, 6))

#summary(HQjust.strain.var_co)

M2.CV <- lme(cV ~ subsp * sex, data=DF.HetC.MixedModel.HQ, random = ~1|strain)

as.data.frame( round(summary(M2.CV)$tTable, 6))

@@

```{r M2.var.Q12, echo=FALSE}
M2.Q12_var_co <- lme(var ~ subsp * sex, data=Q12_mouse_table_w.Ages, random = ~1|strain)

as.data.frame( round(summary(M2.Q12_var_co)$tTable, 6))

M2.Q12_CV <- lme(cV ~ subsp * sex, data=Q12_mouse_table_w.Ages, random = ~1|strain)

as.data.frame( round(summary(M2.Q12_CV)$tTable, 6))
```

$$mouse .var CO ~=~ subsp * sex  + \varepsilon$$
```{r M3.var, echo=FALSE, eval=FALSE}
M3.var <- lm(var ~ subsp * sex * strain, data=DF.HetC.MixedModel.HQ)
as.data.frame( round(summary(M3.var)$tTable, 6))

M3.cv <- lm(cV ~ subsp * sex * strain, data=DF.HetC.MixedModel.HQ)
as.data.frame( round(summary(M3.cv)$tTable, 6))
#these models won't work
```

```{r M3VAR.Q12, echo=FALSE, eval=FALSE}
#ugh, don't know how to get these to work 
M3.Q12_var <- glm(var ~ subsp * sex * strain, data=Q12_mouse_table_w.Ages)

as.data.frame(summary(M3.Q12_var)$tTable, 6)
#results don't show

M3.Q12_cv <- lm(cv ~ subsp * sex * strain, data=Q12_mouse_table_w.Ages)

as.data.frame( round(summary(M3.Q12_cv)$tTable, 6))
```

```{r M4.var, echo=FALSE, eval=FALSE}
M4.var <- lm(var ~ sex * strain, data=DF.HetC.MixedModel.HQ)

as.data.frame( round(summary(M4.var)$tTable, 6))

M4.cv <- lm(cV ~ sex * strain, data=DF.HetC.MixedModel.HQ)

as.data.frame( round(summary(M4.cv)$tTable, 6))
```

```{r M4.var_Q12, echo=FALSE, eval=FALSE}

M4.var_Q12 <- lm(var ~ sex * strain, data=Q12_mouse_table_w.Ages)
as.data.frame( round(summary(M4.var_Q12)$tTable, 6))

#these results not print / working
M4.CV_Q12 <- lm(cV ~ sex * strain, data=Q12_mouse_table_w.Ages)
as.data.frame( round(summary(M4.CV_Q12)$tTable, 6))
```

$$mouse .var CO ~=~ subsp * sex + rand(strain * sex) + \varepsilon$$

#### Summary variance

I couldn't get many of the models to return results.

@@

When Molossinus is coded as a separate subspecies, then interaction terms are no longer significant.  (subsp *sex )  interaction?

>>>>

```{r mixed.model.data.setup, echo=FALSE, warning=FALSE, eval=FALSE}
# MOVE this into BIVDATA SETUP!!!

#should be able to use the loaded 
#DF.HetC.MixedModel.HQ

@@

#afriad this is removing NA's
#DF.HetC.MixedModel.HQ <- DF.HetC.MixedModel[( (DF.HetC.MixedModel$age.weeks < 25) &  #(DF.HetC.MixedModel$age.days != 13) ) | is.na(DF.HetC.MixedModel$age.weeks), ] #keeps mice without ages, and removes juvinilles

@@

```{r MixedModel.1, echo=FALSE, digits = 3}
#THIS IS THE NEW MIXED MODEL
#should be able to use the loaded 
#DF.HetC.MixedModel.HQ

MLH1.MM.M1 <- lmer(mean_co ~ subsp * sex + (1|strain), data=DF.HetC.MixedModel.HQ) 
#summary(Reduced.strain.HQ)

summary(MLH1.MM.M1)
drop1(MLH1.MM.M1, test="Chisq") # p = 0.0002045 ***
#only the interaction term shows up, when the 2 terms are additive, they both show up in the drop1 results
anova(MLH1.MM.M1)

@@

as.data.frame(round(summary(MLH1.MM.M1)$coefficients, 6))
#add a aption to this code chunk
#ranef(Reduced.strain.HQ)
```

@@

## Within Mouse Variance in CO Count per Cell

<!-- intro
in addition to the mean gwRR, the variance is also an important metric to describe

@@

meiotic Checkpoints for ensuring consistency (Hochwagen and Amon 2006, Li and Schimenti 2007, Jaramillo-Lambert 2010 and Kauppi 2013)
-->

Histograms might not be the best way to represent, but they show the general pattern than female mice have higher within mouse variation for CO counts.

```{r variance.hist, echo=FALSE, eval=FALSE}
#figures of histograms,

#DF.HetC.MixedModel.HQ
Var.Hist <- ggplot(DF.HetC.MixedModel.HQ, aes(x = var, fill=sex)) + 
  geom_histogram(binwidth = 1, alpha=.6)  + 
  ggtitle("Histograms of within mouse variance, 10 cell min")
Var.Hist <- Var.Hist   + facet_wrap(~ strain, scales="free")

Var.Hist.15cells <- ggplot(DF.HetC.MixedModel.HQ[DF.HetC.MixedModel.HQ$Ncells >14,], aes(x = var, fill=sex)) + 
  geom_histogram(binwidth = 1, alpha=.6)  +   ggtitle("Histograms of within mouse variance, 15 cell min")
Var.Hist.15cells <- Var.Hist.15cells   + facet_wrap(~ strain, scales="free")

Q12.var.hist.plot <- ggplot(Q12_mouse_table_w.Ages, aes(x = var, fill=sex)) + 
  geom_histogram(binwidth = 1, alpha=.6)  +   ggtitle("Histograms of within mouse variance, Q12")

Q12.var.hist.plot <- Q12.var.hist.plot   + facet_wrap(~ strain, scales="free")
Q12.var.hist.plot

#some LEW female still have v high within mouse variance (2 mice, )

@@

A general pattern is that the female data has higher within mouse variance (both var and cV). Some strains more than others (invetigate potential outliers). (LEW females with top 3 variance are all from different dissection dates)
This pattern also holds for within mouse cv for MLH1 count (not shown).

**todo**

Test if dissection date is more significant predictor than strain (for females) (Effect doc)
Also test how the number of 0CO bivalents per cell affects mouse with variance. (lm)

@@

#DF.HetC.MixedModel.HQ
cV.Hist <- ggplot(DF.HetC.MixedModel.HQ, aes(x = cV, fill=sex)) + 
  geom_histogram(binwidth = 1, alpha=.6)  + 
  ggtitle("cV of within mouse variance, 10 cell min")
cV.Hist <- cV.Hist   + facet_wrap(~ strain, scales="free")
cV.Hist
```

### Modeling Within Mouse Variance

XXX (old 1.)	HetC.MixedModel.HQ <- lme(mouse.metric ~ subsp*sex, data=DF.HetC.MixedModel.HQ, random=list(strain=pdDiag(~sex) ) )

1.(new 1)	Reduced.strain.HQ <- lmer(mouse.metric ~ subsp * sex+ (1|strain), data=DF.HetC.MixedModel.HQ)

3.	lm(mouse.metric ~ subsp \* sex \* strain, data=DF.HetC.MixedModel.HQ)

4.	lm(mouse.metric ~ sex * strain, data=DF.HetC.MixedModel.HQ)

$$mouse metric  ~=~ subsp * sex + rand(strain)+ \varepsilon$$

```{r M1.MLH1.var_cv, echo=FALSE}
#modeling within mouse variance
#use  DF.HetC.MixedModel.HQ

#HQ var.CO
MLH1var_M1 <- lmer( var ~ subsp * sex +(1|strain), data=DF.HetC.MixedModel.HQ )
summary(MLH1var_M1)

drop1(MLH1var_M1, test="Chisq") #interaction term NS
#when switched to additive fixed, sex is very sig

anova(MLH1var_M1)#sex explains most variance...

exactRLRT(MLH1var_M1) #strain
as.data.frame( round(summary(MLH1var_M1)$tTable, 6))

#CV
MLH1cv_M1 <- lmer( cV ~ subsp * sex +(1|strain), data=DF.HetC.MixedModel.HQ )
summary(MLH1cv_M1)

drop1(MLH1cv_M1, test="Chisq") #interaction term NS
#when switched to additive fixed, sex is very sig
anova(MLH1cv_M1)  #sex F value is large
exactRLRT(MLH1cv_M1) #strain

as.data.frame( round(summary(M1.cv_co)$tTable, 6))

#USE Q12 (Q123) table
#Q12_mouse_table_w.Ages
#Q12 var
MLH1var_M1_Q12 <- lmer( var ~ subsp * sex +(1|strain), data=Q12_mouse_table_w.Ages )

drop1(MLH1var_M1_Q12, test="Chisq")#sig
anova(MLH1var_M1_Q12)
exactRLRT(MLH1var_M1_Q12)#NS

as.data.frame( round(summary(MLH1var_M1_Q12)$coefficients, 6))

#Q12 cv
MLH1cv_M1_Q12 <- lmer( cV ~ subsp * sex +(1|strain), data=Q12_mouse_table_w.Ages )

drop1(MLH1cv_M1_Q12, test="Chisq")#NS
anova(MLH1cv_M1_Q12)#sex large F
exactRLRT(MLH1cv_M1_Q12)#NS

as.data.frame( round(summary(MLH1cv_M1_Q12)$coefficients, 6))

@@

The fixed sex effect has the largest effect sizes. All are negative (for male).  (I'm not sure if I should switch to testing the additive fixed effects by themselves) - I'm pretty sure the outcome would be that the sex effect gets very large in those models.
The strain effects are only slightly significant for within mouse variance measures in the full dataset (includes low quality cells)

$$mouse CO metric  ~=~ subsp * sex * strain + \varepsilon$$
M2 slash M3

```{r M3.MLH1.var, echo=FALSE}

MLH1var_M3 <- lm( var ~ subsp * sex * strain,  data=DF.HetC.MixedModel.HQ )

summary(MLH1var_M3)
anova(MLH1var_M3)  
as.data.frame( round(summary(MLH1var_M3)$tTable, 6))

MLH1cv_M3 <- lm(cV ~ subsp*sex*strain, data=DF.HetC.MixedModel.HQ)

summary(MLH1cv_M3)

anova(MLH1cv_M3)
as.data.frame( round(summary(M1.cv_co)$tTable, 6))

#Q12 var
MLH1var_M3_Q12 <- lm( var ~ subsp*sex*strain,  data=Q12_mouse_table_w.Ages )

summary(MLH1var_M3_Q12)
anova(MLH1var_M3_Q12)  
as.data.frame( round(summary(MLH1var_M3_Q12)$tTable, 6))

#Q12 cv
MLH1cv_M3_Q12 <- lm( cV ~ subsp*sex*strain, data=Q12_mouse_table_w.Ages )

summary(MLH1cv_M3_Q12)
anova(MLH1cv_M3_Q12)#sex large F

as.data.frame( round(summary(MLH1cv_M3_Q12)$coefficients, 6))

@@

For LEW - females must have high variance,  for CV Lew strain alone has increased cv. Consider plotting the variance - and diving deeper into this pattern
Most of the strain specific differences in within mouse variance go away in the Q12 dataset.

$$mouse CO metric  ~=~ subsp * sex * strain + \varepsilon$$

```{r M4.MLH1.var, echo=FALSE}

MLH1var_M4 <- lm( var ~ sex * strain,  data=DF.HetC.MixedModel.HQ )

summary(MLH1var_M4)

drop1(MLH1var_M4, test="Chisq")
anova(MLH1var_M4)
as.data.frame( round(summary(MLH1var_M4)$tTable, 6))

MLH1cv_M4 <- lm(cV ~ sex * strain, data=DF.HetC.MixedModel.HQ)

summary(MLH1cv_M4)

drop1(MLH1cv_M4, test="Chisq")
anova(MLH1cv_M4)  
exactRLRT(MLH1cv_M4)
as.data.frame( round(summary(MLH1cv_M4)$tTable, 6))

### Q12 var
MLH1var_M4_Q12 <- lm( var ~ subsp*sex*strain, data=Q12_mouse_table_w.Ages )
anova(MLH1var_M4_Q12)
as.data.frame( round(summary(MLH1var_M4_Q12)$coefficients, 6))

#Q12 cv
MLH1cv_M4_Q12 <- lm( cV ~ subsp*sex*strain, data=Q12_mouse_table_w.Ages )

summary(MLH1cv_M4_Q12)
anova(MLH1cv_M4_Q12)#sex large F

as.data.frame( round(summary(MLH1cv_M4_Q12)$coefficients, 6))

@@

#### Summary witin mouse Variance

@@

(some strains have significant effect -- some LEW females had large leverage).
